### 2021-10-14 paper/paper.tex
changes

\newcommand{\dist}[2]{d_{{#1},{#2}}}
\newcommand{\level}[1]{\texttt{level}({#1})}

@@

%\newcommand{\fixme}[1]{\noindent{\color{red}\textbf{FIXME:}  {#1}}}
\newcommand{\fixme}[1]{}
\newcommand{\fixmemike}[1]{\noindent{\color{blue}\textbf{FIXME (Mike):}  {#1}}}

@@

\section{The Tree Loss}
\label{sec:tree}

This section first formally defines our problem setting,
then formally defines the tree loss.
We conclude with a comparison between the tree loss and related loss functions.

\subsection{Problem Setting}

@@

\subsection{The Tree Loss}

The tree loss reparameterizes the weight matrix used in the cross entropy loss in order to enforce that semantically similar labels will have similar weight vectors.
The reparameterization is defined according to a \emph{label tree} structure that captures the similarity between labels.
And we propose two variants of the tree loss called the $U$ and $V$ variants that use slightly different label tree structures.
The main purpose of the $U$-tree structure is to simplify the analysis and exposition of the $V$-tree structure,
and we show later in our theoretical results that the $V$-tree is strictly better than the $U$-tree.
Figure \ref{fig:labeltree} shows examples of both.

\subsubsection{The $U$-Tree Loss}

In a $U$-tree, every node is a class and the class of all internal nodes matches the class of one of its children.
For each class $i$, we define $\parent{i}$ to be the parent of the upper-most node for class $i$,
and the $\texttt{parent}$ of the root is the 0 vector..
For example, in the example of Figure \ref{fig:labeltree},
we have
\begin{equation*}
\begin{split}
    \parent{\texttt{husky}} &= \texttt{boxer} \text{, and}\\
    \parent{\texttt{bear}} &= \texttt{boxer} \text{, and}\\
    \parent{\texttt{boxer}} &= \texttt{truck} .\\
\end{split}
\end{equation*}
We associate for each class $i$ a new parameter vector $\uu_i$ and set
\begin{equation}
    \w_i = \w_{\parent{i}} - \uu_i,
    \label{eq:uu}
\end{equation}
and we define the parameter matrix $U : \mathbb R^{k\times d}$ to be $(\uu_1;...;\uu_k)$.
The $U$-tree loss is then defined by substituting \eqref{eq:uu} into \eqref{eq:xentropy} to get
\begin{equation*}
    \ell(U;(\x,y)) = - \log \frac {\exp(-(\trans\uu_{\parent{y}}-\trans\uu_y) \x)}{\sum_{j=1}^k \exp(-(\trans\uu_{\parent{j}}-\trans \uu_j) \x)}
\end{equation*}
We use the same $\ell$ notation for the standard cross entropy loss function and the $U$-Tree loss because the function is exactly the same;
the only difference is how we represent the parameters.

\subsubsection{The $V$-Tree Loss}

The $V$-tree is constructed from the $U$-tree by replacing all non-leaf nodes with ``pseudoclasses''.
Let $k'$ be the total number of nodes in the $V$-tree;
so $k'$ equals $k$ plus the number of internal nodes.
For each class $i$, we let the sequence $P_i$ denote the path from the leaf node to the root.
For example, using the $V$-tree in Figure \ref{fig:labeltree},
we have
\begin{equation*}
\begin{split}
    P_{\texttt{sheepdog}} &= (\texttt{sheepdog}, \textit{pseudo3}, \textit{pseudo2}, \textit{pseudo1} ) \\
    P_{\texttt{truck}} &= (\texttt{truck}, \textit{pseudo4}, \textit{pseudo1} )
\end{split}
\end{equation*}
Next, we associate a vector $\vv_i$ to each node $i$ in the tree structure.
In particular, there is a vector $\vv_i$ for each class and for each pseudo-class,
so there are more parameter vectors for the $V$-tree loss than either the $U$-tree loss or the standard cross entropy loss.

Finally, we reparameterize the weights for the cross entropy loss as the sum of the vectors along the paths.
Specifically, we set the weight vector for class $i$ to be 
\begin{equation}
    \label{eq:covertree-reparam}
    \w_j = \sum_{j\in P_i} \vv_i
    .
\end{equation}
The $V$-tree loss is defined by substituting \eqref{eq:covertree-reparam} into \eqref{eq:xentropy} to get
\begin{equation}
    \ell(V;(\x,y)) = - \log \frac {\exp(-\sum_{k\in P_y}\trans\vv_k \x)}{\sum_{j=1}^k \exp(- \sum_{k\in P_j}\trans\vv_k \x)}
    .
\end{equation}
We continue to use the $\ell$ function to represent both the standard cross entropy loss and the $V$-tree loss function to emphasize that these are the same loss functions,
just with different parameterizations of the weight matrix.

@@

@@

Both the $U$-tree and $V$-tree losses are easy to implement in practice using the built-in cross entropy loss function in libraries like PyTorch \citep{NEURIPS2019_9015} or Tensorflow \citep{tensorflow2015-whitepaper}.
The only change needed is to define the loss in terms of the sum of vectors $V$ instead of a variable tensor $W$.
This ensures that the loss is implemented efficiently and is numerically stable.

In practice, the $V$-tree loss is slightly slower than the standard cross entropy loss and $U$-tree losses due to the additional summation over the paths.
In our experiments, we found this this to cause the $V$-tree loss to be about twice as slow.

@@

It is well known that when using the standard cross entropy loss,
SGD converges at a rate of $O(\sqrt{k/n})$.
Our theoretical results asymptotically improve this convergence rate.

With an appropriately constructed label tree,
we show convergence at a rate of $O(\sqrt{\log k/n})$ or $O(\sqrt{k^{1-1/c}/n})$,
where $c$ is a measure of difficulty of the problem.
But what happens if the tree is poorly constructed?
Remarkably, nothing bad.
We also show that for arbitrarily bad label tree structures,
the $U$ and $V$ losses maintain a $O(\sqrt{k/n})$ convergence rate.
This suggests that there is essentially no downside to using the tree loss parameterizations.

\fixme{
Our analysis relies on simple properties of stochastic gradient descent,
and we view the simplicity of our analysis as a major strength of the tree loss.
In this section, we first review standard results of stochastic gradient descent needed for our analysis.
Then we formally state and prove our main result.
}

@@

SGD is a standard algorithm for optimizing convex losses,
and it has many variations.
Our analysis uses the results from Chapter 14 of \citet{shalev2014understanding} which are based off of a version of SGD that uses parameter averaging.\footnote{
Other variants of SGD have similar results,
and our results are easy to generalize to other SGD variants.
We refer the reader to \citet{shalev2014understanding} for a discussion of these variants.}
In this section we recall these results,
occasionally changing the names of variables to avoid conflicts with variable names used elsewhere in this paper.

We assume we are optimizing a function $f : \Theta \to \mathbb R$,
and that we are only given noisy samples of the function and its gradient.
We let $\theta^{t}$ denote the parameter value at iteration $t$,
and $\eta : \mathbb R$ be the step size.
Then the parameter update for step $t$ is defined to be
\begin{equation}
    \theta^{(t+1)} = \theta^{(t)} - \eta \nabla f(\theta^{(t)}).
\end{equation}
When using parameter averaging, the final model parameters are defined to be
\begin{equation}
    \bar\theta = \frac1T \sum_{i=1}^T \theta^{(t)}
    .
\end{equation}
Parameter averaging is not often used in practice because of the memory requirements,
but we use it here due because theorems using parameter averaging are typically much easier to state and prove.
Equivalent results without parameter averaging typically require and additional $\log T$ factor in their convergence rates\citep{}.

@@

\begin{theorem}
    \label{thm:sgd}
    Let $\beta,\rho>0$.
    Let $f$ be a convex function and let $\star \theta = \argmin_{\theta : \ltwo{\theta}\le \beta} f(\theta)$.
    Assume that SGD is run for $T$ iterations with $\eta = \sqrt{\frac{\beta^2}{\rho^2 T}}$.
    Assume that for all $t$, $\ltwo{\nabla f(\theta^{(t)})} \le \rho$.
    Then,
    \begin{equation}
        \E f(\bar \theta) - f(\star \theta) \le \frac{\beta\rho}{\sqrt{T}}
        .
    \end{equation}
\end{theorem}

@@

We now introduce more statistical learning theory notation,
then apply Theorem \ref{thm:sgd} above to bound the rate of learning.

Define the true loss of our learning problem as
\begin{equation}
    L_D(A) = \E_{(\x,y)\sim D} \ell(A; (\x, y))
\end{equation}
where $A \in \{W, U, V\}$ is a parameterization of the cross entropy loss.
The optimal parameter vector is defined to be the minimum of the true loss:
\begin{equation}
    \star A = \argmin_{A} L_D(A).
\end{equation}
In order to apply Theorem \ref{thm:sgd} to a learning problem,
we make the following assumption about our the size of our data points:
\begin{assumption}
    \label{ass:lip}
    For each feature vector $\x$ in the data set, 
        $\ltwo{\x} \le \rho$
        .
\end{assumption}
Under this assumption, the cross entropy loss $\ell$ is $\rho$-Lipschitz for all parameterizations,
and the true loss $L_D$ is therefore also $\rho$-Lipschitz.

Applying Theorem \ref{thm:sgd} with $f=L_D$ gives
\begin{corollary}
    \label{ref:cor:A}
    Under Assumption \ref{ass:lip}, we have that for any parameterization $A$ of the cross entropy loss,
    \begin{equation}
        \E L_D(\bar A) - L_D(\star A) \le \frac{\lF{\star A}\rho}{\sqrt n}.
        \label{eq:Aconv}
    \end{equation}
\end{corollary}

We can now bound the convergence of the standard cross entropy loss and the $U$ and $V$-tree losses by bounding the Frobenius norms of their parameter matrices.

In order to recover the standard convergence rate of multiclass classification, we make the following assumption.
\begin{assumption}
    \label{ass:B}
    For each class $i$, the optimal parameter vector $\star\w_i$ satisfies
        $\ltwo{\star\w_i} \le B$
        .
\end{assumption}
Note that this assumption is not limiting because the value of $B$ need not be known during the optimization procedure.

It follows that
\begin{equation}
    \lF{\star W}^2 = \sum_{i=1}^k \ltwo{\star\w_i} \le kB^2.
    \label{eq:starW}
\end{equation}
Substituting Eq \eqref{eq:starW} into \eqref{eq:Aconv} gives the following bound.

\begin{theorem}
\label{theorem:xentropy}
    Under assumptions \ref{ass:lip} and \ref{ass:B} above,
then the parameter vector $\bar W$ estimated from one epoch of SGD satisfies
\begin{equation}
    \E L_D(\bar W) - L_D(W^*)
    \le \frac {\sqrt kB\rho}{\sqrt n}
    .
\end{equation}
\end{theorem}
\fixme{
It is important to note that this result above is an upper bound and not a lower bound.
This $O(\sqrt{k/n})$ convergence rate matches the convergence rate implied by the Nataranj dimension \citep{}.
We are not aware of any lower bounds for SGD in the Lipschitz setting.
\citet{nguyen2018tight,jentzen2020lower} study the strongly convex case and find that the upper bounds match the lower bounds.
}

\subsection {Bad Tree Structures}

\begin{lemma}
    \label{lemma:starV}
    We have that for all $V$-trees derived from a $U$-tree,
    \begin{equation}
        \lF{\star V} \le \lF{\star U}.
    \end{equation}
\end{lemma}
\fixme{
\begin{proof}
\end{proof}
}

The following lemma shows that even for arbitrarily bad tree structures,
the bound on the parameter spaces cannot be too bad.

\begin{lemma}
    \label{lemma:starU}
    Under assumption \ref{ass:B},
    the following bound holds for all $U$-tree structures:
    \begin{equation}
        \lF{\star U} \le 2\sqrt{k}B.
    \end{equation}
\end{lemma}
\begin{proof}
    We have
    \begin{align}
        \lF{\star U}^2 
        &= \sum_{i=1}^k \ltwo{\star\uu_i}^2 \\
        &= \sum_{i=1}^k \ltwo{\star\w_i - \star\w_{\parent{i}}}^2 \\
        &\le \sum_{i=1}^k \left(\ltwo{\star\w_i} + \ltwo{\star\w_{\parent{i}}} \right)^2 \\
        &\le \sum_{i=1}^k (2B)^2 \\
        &\le k (2B)^2
        .
    \end{align}
\end{proof}

\subsection{Good Tree Structures}

\begin{assumption}
We are given a distance metric $d$ over the labels such that for all labels $i$ and $j$,
\begin{equation}
    \frac 1 \lambda d(i,j)
    \le \ltwo{\star \w_i - \star \w_j}
    \le \lambda d(i, j).
\end{equation}
We denote by $c$ the doubling dimension of the metric $d$.
\end{assumption}

\begin{lemma}
    \label{lemma:main}
    When $c\le1$, we have that
    \begin{equation}
        \lF{\star U} \le \tfrac{1}{\sqrt2}\lambda B \sqrt{\log_2 k}.
        \label{eq:c<=1}
    \end{equation}
    When $c>1$, we have that
    \begin{equation}
        \lF{\star U} \le \sqrt{5}\lambda B \sqrt{k^{(1-1/c)}}.
        \label{eq:c>1}
    \end{equation}
\end{lemma}

\fixme{
\begin{proof}
    First, we review some simple properties of the cover tree structure that will be needed in our proof.
    Let $q$ denote the maximum number of children of any node in the cover tree.
    A standard property of cover trees is that $q \le (4^c)$ \citep[Lemma 25]{izbickithesis}.
    Let $Q_i$ denote the number of nodes at level $i$ in the cover tree.
    Then we have that $|Q_i| \le q^i.$
    Finally, let $r$ be the height of the cover tree.
    Without loss of generality, we analyze the bound when the cover tree is perfectly balanced and so $r = \log_q k = \log_2 k / (2c)$.

When $c>$, then $q>4$.
We first show the following equality.
\begin{align}
\left(\frac q 4\right)^r
&=
\left(\frac q 4\right)^{\log_q k}\\
&=
\frac{k}{4^{\log_q k}}\\
&=
\frac{k}{4^{{\log_4 k}/{\log_4 q}}}\\
&=
\frac{k}{k^{1/\log_4 q}}\\
&=
k^{\left(1 - {1}/{\log_4 q}\right)} \\
&=
k^{\left(1 - 1/c\right)}
.
\label{eq:lFU:frac}
\end{align}
Substituting into Eq \eqref{eq:precases}, we have
\begin{align}
    \lF{\star U}^2
    &= \lambda^2B^2 \left(\frac{(q/4)^{r+1}-1}{q/4-1}\right) \\
    \label{eq:lFU:4}
    &= \lambda^2B^2 \left(\frac{(q/4)k^{(1-1/c)}-1}{q/4-1}\right) \\
    &\le \lambda^2B^2 \left(\frac{(q/4)k^{(1-1/c)}}{q/4-1}\right) \\
    &\le 5 \lambda^2B^2 k^{(1-1/c)}
    \label{eq:lFU:5}
\end{align}
    Inequality \eqref{eq:lFU:5} follows because $\tfrac{q/4}{q/4-1} \le 5$ for integer values of $q>4$.
\end{proof}
}

@@

}
\end{document}

>>>>

\newcommand{\dist}[2]{d_{{#1},{#2}}}
\newcommand{\level}[1]{\texttt{level}({#1})}
\newcommand{\depth}[1]{\texttt{depth}({#1})}

@@

\newcommand{\fixme}[1]{\noindent{\color{red}\textbf{FIXME:}  {#1}}}
%\newcommand{\fixme}[1]{}
\newcommand{\fixmemike}[1]{\noindent{\color{blue}\textbf{FIXME (Mike):}  {#1}}}

@@

\section{Problem Setting}
\label{sec:tree}

@@

The cross entropy loss has no constraints on the weight matrix that cause similar classes to have similar parameter vectors.
The tree loss fixes this problem, resulting in faster convergence and better generalization.

\section{The Tree Loss}

The main idea of the tree loss is that similar classes should be forced to ``share'' some of their parameters.
%This sharing causes parameter updates in SGD to also update the parameters of similar classes,
%resulting in improved generalization.
The tree loss refactors the cross entropy loss's weight matrix in order to enforce this sharing.

We propose two variants of the tree loss which enforce this sharing in different ways.
We begin by introducing the $U$-tree loss,
which is simpler to explain and analyze.
Then, we introduce the $V$-tree loss,
which improves on the $U$-tree loss and is the loss we suggest using in practice.
Whenever we refer to the tree loss without a qualifier,
we mean the $V$-tree loss.

%Our analysis in Section \ref{sec:theory} will require a metric structure over the class labels.
%This metric structure is not needed to explain the construction of the tree loss itself,
%however, so we defer this explanation until later.

%The tree loss reparameterizes the weight matrix used in the cross entropy loss in order to enforce that semantically similar labels will have similar weight vectors.
%The reparameterization is defined according to a \emph{label tree} structure that captures the similarity between labels.
%And we propose two variants of the tree loss called the $U$ and $V$ variants that use slightly different label tree structures.
%The main purpose of the $U$-tree structure is to simplify the analysis and exposition of the $V$-tree structure,
%and we show later in our theoretical results that the $V$-tree is strictly better than the $U$-tree.
%Figure \ref{fig:labeltree} shows examples of both.

\subsection{The $U$-Tree Loss}

Explaining our new $U$-tree parameterization requires introducing some notation.

We define a $U$-tree over the class labels to be a tree where each leaf node is represented by a label,
and each internal node shares a label with a child node.
Figure \ref{fig:labeltree} (left) shows an example $U$-tree structure.
The definition of a $U$-tree is motivated by the cover tree data structure \citep{beygelzimer2006cover},
which generates $U$-tree structures given a metric over the class labels.

%In a $U$-tree, every node is a class and the class of all internal nodes matches the class of one of its children.
For each class $i$, we let the sequence $P_i$ denote the path from the leaf node to the root with duplicates removed.
For example, using the $U$-tree in Figure \ref{fig:labeltree},
we have
\begin{equation*}
\begin{split}
    P_{\texttt{bear}} &= (\texttt{bear}, \texttt{boxer}, \texttt{truck}) \\
    P_{\texttt{sheepdog}} &= (\texttt{sheepdog}, \texttt{boxer}, \texttt{truck}) \\
    P_{\texttt{truck}} &= (\texttt{truck})
\end{split}
\end{equation*}
For each class $i$, we define $\parent{i}$ to be the first node in $P_i$ not equal to $i$.
In other words, $\parent{i}$ is the parent of the upper-most node for class $i$.
%For example, using the $U$-tree structure in Figure \ref{fig:labeltree},
%we have
%\begin{equation*}
%\begin{split}
    %\parent{\texttt{husky}} &= \texttt{boxer} \text{, and}\\
    %\parent{\texttt{bear}} &= \texttt{boxer} \text{, and}\\
    %\parent{\texttt{boxer}} &= \texttt{truck} .\\
%\end{split}
%\end{equation*}

We are now ready to introduce the $U$-tree parameterization.
Intuitively, we will represent each parameter as the difference between itself and it's parent in the $U$-tree.
We associate for each class $i$ a new parameter vector $\uu_i$ defined to be
\begin{equation}
    \uu_i =
    \begin{cases}
        \w_i                    & \text{if $i$ is the root} \\
        \w_{\parent{i}} - \w_i  & \text{otherwise}
    \end{cases}
    ,
\end{equation}
and we define the parameter matrix $U : \mathbb R^{k\times d}$ to be $(\uu_1;\uu_2;...;\uu_k)$.
We can recursively rewrite the original parameter vectors in terms of this new parameterization as
\begin{equation}
    \w_i = \sum_{j\in P_i} \uu_i
    .
    \label{eq:uu}
\end{equation}
The $U$-tree loss is then defined by substituting \eqref{eq:uu} into \eqref{eq:xentropy} to get
\begin{equation*}
    %\ell(U;(\x,y)) = - \log \frac {\exp(-(\trans\uu_{\parent{y}}-\trans\uu_y) \x)}{\sum_{j=1}^k \exp(-(\trans\uu_{\parent{j}}-\trans \uu_j) \x)}
\ell(U;(\x,y)) = - \log \frac {\exp(-\sum_{j\in P_y}\trans\uu_j \x)}{\sum_{i=1}^k \exp(- \sum_{j\in P_i}\trans\uu_j \x)}
\end{equation*}
We use the same $\ell$ notation for the standard cross entropy loss function and the $U$-Tree loss because the function is exactly the same;
the only difference is how we represent the parameters during the training procedure.

\subsection{The $V$-Tree Loss}

The $V$-tree is constructed from the $U$-tree by replacing all non-leaf nodes with ``pseudoclasses''.
Let $k'$ denote the total number of classes plus the total number of pseudoclasses.
%so $k'$ equals $k$ plus the number of internal nodes.
%For each class $i$, we let the sequence $P_i$ denote the path from the leaf node to the root.
%For example, using the $V$-tree in Figure \ref{fig:labeltree},
%we have

We let $P_i$ be defined as the path from leaf node $i$ to the root as before,
but note that there are now no duplicates to remove and so the paths are longer.
For example,
\begin{equation*}
\begin{split}
    P_{\texttt{bear}} &= (\texttt{bear}, \textit{pseudo2}, \textit{pseudo1}) \\
    P_{\texttt{sheepdog}} &= (\texttt{sheepdog}, \textit{pseudo3}, \textit{pseudo2}, \textit{pseudo1} ) \\
    P_{\texttt{truck}} &= (\texttt{truck}, \textit{pseudo4}, \textit{pseudo1} )
\end{split}
\end{equation*}
%Next, we associate a vector $\vv_i$ to each node $i$ in the tree structure.
%In particular, there is a vector $\vv_i$ for each class and for each pseudo-class,
%so there are more parameter vectors for the $V$-tree loss than either the $U$-tree loss or the standard cross entropy loss.
The $\vv_i$ and $V$ parameters are now defined analogously to the $\uu_i$ and $U$ parameters, but over the $V$-tree structure instead of the $U$-tree structure.
An important distinction between the $V$ and $U$ parameter matrices is that they will have different shapes.
The $U$ matrix has shape $k \times d$, which is the same as the $W$ matrix,
but the $V$ matrix has shape $k' \times d$,
which is potentially a factor of up to 2 times larger.

%Finally, we reparameterize the weights for the cross entropy loss as the sum of the vectors along the paths.
%Specifically, we set the weight vector for class $i$ to be 
%\begin{equation}
    %\label{eq:covertree-reparam}
    %\w_j = \sum_{j\in P_i} \vv_i
    %.
%\end{equation}
%The $V$-tree loss is defined by substituting \eqref{eq:covertree-reparam} into \eqref{eq:xentropy} to get
The $V$-tree loss is defined by using the $V$ matrix to parameterize the cross entropy loss, giving
\begin{equation}
    \ell(V;(\x,y)) = - \log \frac {\exp(-\sum_{k\in P_y}\trans\vv_k \x)}{\sum_{j=1}^k \exp(- \sum_{k\in P_j}\trans\vv_k \x)}
    .
\end{equation}
We continue to use the $\ell$ function to represent both the standard cross entropy loss and the $V$-tree loss function to emphasize that these are the same loss functions,
just with different parameterizations of the weight matrix.

@@

This reparameterization is reminiscent of the way that fastText \citep{bojanowski2017enriching} reparameterizes the word2vec \citep{Mikolov2013EfficientEO} model to improve statistical efficiency.
Two notable difference, however, are that fastText is a domain-specific adaptation and provides no theoretical guarantees;
our tree loss works on any domain and provides theoretical guarantees.

@@

Both the $U$-tree and $V$-tree losses are easy to implement in practice using the built-in cross entropy loss function in libraries like PyTorch \citep{NEURIPS2019_9015} or Tensorflow \citep{tensorflow2015-whitepaper}.
%The only change needed is to pass in a variable representing the $U$/$V$ parameter matrices to the cross entropy loss function
The only change needed is to represent the $W$ parameter matrix in your code as an appropriate sum over the $U$ or $V$ parameter matrices.
The magic of automatic differentiation will then take care of the rest and ensure that the underlying $U$ or $V$ matrix is optimized.
Our TreeLoss library\footnote{URL blinded for peer review.} provides easy-to-use functions for generating these matrices,
and so modifying code to work with the tree loss is a 1 line code change.

In practice, the tree loss is slightly slower than the standard cross entropy loss due to the additional summation over the paths.
In our experiments, we observed an approximately 2x slowdown.

@@

It is well known that when using the standard cross entropy loss,
SGD converges at a rate of $O(\sqrt{k/n})$.
The tree loss's parameter sharing scheme asymptotically improves this convergence rate.

We assume that the class labels have a metric structure and let $c$ denote the doubling dimension of this metric.
This $c$ is a measure of the difficulty of our problem,
with larger $c$ implying more complex metrics and a more difficult learning problem.
This is captured in our convergence rates,
which show show that the tree loss converges at a rate of $O(\sqrt{\log k/n})$ for $c\le1$ or $O(\sqrt{k^{1-1/c}/n})$ for $c>1$.

But what happens if the tree is poorly constructed?
Remarkably, nothing bad.
We also show that for arbitrarily bad label tree structures,
the $U$ and $V$ losses maintain a $O(\sqrt{k/n})$ convergence rate.
This suggests that there is essentially no downside to using the tree loss parameterizations.

Our analysis relies on standard properties of stochastic gradient descent.
In this section, we first review these results,
then we formally state and prove our main result.

@@

SGD is a standard algorithm for optimizing convex losses,
and it has many variations.
Our analysis uses the results from Chapter 14 of \citet{shalev2014understanding} which are based off of a version of SGD that uses parameter averaging.\footnote{
Other variants of SGD have similar results,
and we refer the reader to \citet{shalev2014understanding} for a discussion of these variants.}
In this section we review these results.
We follow the notation of \citet{shalev2014understanding} as closely as possible,
but we occasionally change the names of variables to avoid conflicts with variable names used elsewhere in this paper.\footnote{
    Specifically, we replace all Latin variables with Greek alternatives.
    We use $\theta$ instead of $\w$, $\nu$ instead of $\vv$, and $\beta$ instead of $B$.
    The Greek letters $\rho$ and $\eta$ have the same meaning in both documents.
}

SGD is an iterative procedure for optimizing a function $f : \Theta \to \mathbb R$ given only noisy samples of the function's gradient.
We let $\theta^{(t)}$ denote the parameter value at iteration $t$,
and $\eta : \mathbb R$ be the step size.
Let $\nu_t$ be a noisy gradient of $f$ satisfying
\begin{equation}
    \E[\nu_t | \theta^{(t)}] = \nabla f(\theta^{(t)}).
\end{equation}
Then the parameter update for step $t$ is defined to be
\begin{equation}
    \theta^{(t+1)} = \theta^{(t)} - \eta \nabla \nu_t
    .
\end{equation}
When using parameter averaging, the final model parameters are defined to be
\begin{equation}
    \bar\theta = \frac1T \sum_{i=1}^T \theta^{(t)}
    .
\end{equation}
Parameter averaging is not often used in practice because it has a large memory overhead,
but it is widely used in theoretical analysis of SGD.
Theorems using parameter averaging are typically much easier to state and prove,
and equivalent results without parameter averaging typically require an additional $\log T$ factor in their convergence rates \citep{rakhlin2011making}.

@@

\begin{theorem}
    \label{thm:sgd}
    Let $\beta,\rho>0$.
    Let $f$ be a convex, $\rho$-Lipschitz function and let $\star \theta = \argmin_{\theta : \ltwo{\theta}\le \beta} f(\theta)$.
    Assume that SGD is run for $T$ iterations with $\eta = \sqrt{\frac{\beta^2}{\rho^2 T}}$.
    %Assume that for all $t$, $\ltwo{\nabla f(\theta^{(t)})} \le \rho$.
    Then,
    \begin{equation}
        \E f(\bar \theta) - f(\star \theta) \le \frac{\beta\rho}{\sqrt{T}}
        .
    \end{equation}
\end{theorem}

@@

We now introduce more statistical learning theory notation,
and apply Theorem \ref{thm:sgd} above to bound the rate of learning.

Define the true loss of our learning problem as
\begin{equation}
    L_D(A) = \E_{(\x,y)\sim D} \ell(A; (\x, y))
\end{equation}
where $A \in \{W, U, V\}$ is a parameterization of the cross entropy loss.
The optimal parameter vector is defined to be the minimum of the true loss:
\begin{equation}
    \star A = \argmin_{A} L_D(A).
\end{equation}
In order to apply Theorem \ref{thm:sgd} to a learning problem,
we make the following assumption about the size of our data points.
\begin{assumption}
    \label{ass:lip}
    For each feature vector $\x$ in the data set, 
        $\ltwo{\x} \le \rho$
        .
\end{assumption}
%For standard data distributions (e.g. any sub-normal distribution),
%$\rho$ is upper bounded by $O(\sqrt{d})$.
Under this assumption, the cross entropy loss $\ell$ is $\rho$-Lipschitz for all parameterizations,
and the true loss $L_D$ is therefore also $\rho$-Lipschitz.

We use SGD to optimize $L_D$.
We let $f=L_D$, and $\theta = A$,
so $\nu_t$ is the gradient of $\ell$ on the $t$th data point.
Applying Theorem \ref{thm:sgd} gives the following convergence bound.
\begin{lemma}
    \label{ref:cor:A}
    Under Assumption \ref{ass:lip},
    we have that for any parameterization $A$ of the cross entropy loss,
    SGD converges at the rate
    \begin{equation}
        \E L_D(\bar A) - L_D(\star A) \le \frac{\lF{\star A}\rho}{\sqrt n}.
        \label{eq:Aconv}
    \end{equation}
\end{lemma}

We emphasize that this convergence rate above is both computational and statistical.
The expression on the left hand side of \eqref{eq:Aconv} is equivalent to the expected generalization error of the parameter $\bar A$,
and so we use the terms ``convergence rate'' and ``generalization error'' essentially interchangeably.

The important intuition of Lemma \ref{ref:cor:A} is that the generalization error of the cross entropy loss depends on the Frobenious norm of the optimal parameter matrix.
We can therefore compare the convergence of the standard parameterization with the $U$/$V$-tree parameterizations by comparing these norms.

Our next task is to recover the standard convergence rate of multi-class classification, and so we will now bound $\lF{\star W}$.
We make the following assumption.
\begin{assumption}
    \label{ass:B}
    For each class $i$, the optimal parameter vector $\star\w_i$ satisfies
        $\ltwo{\star\w_i} \le B$
        .
\end{assumption}
%Note that this assumption is not limiting because the value of $B$ need not be known during the optimization procedure.
%
It follows that
\begin{equation}
    \lF{\star W}^2 = \sum_{i=1}^k \ltwo{\star\w_i} \le kB^2.
    \label{eq:starW}
\end{equation}
Substituting Eq \eqref{eq:starW} into \eqref{eq:Aconv} gives the following bound.
\begin{corollary}
\label{theorem:xentropy}
    Under assumptions \ref{ass:lip} and \ref{ass:B},
    then the generalization error of the standard cross entropy parameterization when trained with SGD is
\begin{equation}
    \E L_D(\bar W) - L_D(W^*)
    \le \frac {\sqrt kB\rho}{\sqrt n}
    .
\end{equation}
\end{corollary}

\ignore{
It is important to note that this result above is an upper bound and not a lower bound.
This $O(\sqrt{k/n})$ convergence rate matches the convergence rate implied by the Nataranj dimension \citep{}.
We are not aware of any lower bounds for SGD in the Lipschitz setting.
\citet{nguyen2018tight,jentzen2020lower} study the strongly convex case and find that the upper bounds match the lower bounds.
}

\subsection{Bounding the Tree Loss}

We now bound the Frobenius norm of $\star U$ in order to bound the generalization of the $U$-tree loss.
By construction, $\lF{\star V} \le \lF{\star U}$,
so this will also bound the convergence rates of the $V$-tree loss.

The analysis is divided into two parts.
First, we make no assumptions that the $U$-tree structure is good,
and we show that even in the worst case $\lF{\star U} = O(\lF{\star W})$.
This implies that using the tree loss cannot significantly hurt our performance.
\begin{lemma}
    \label{lemma:starU}
    Under assumption \ref{ass:B},
    the following bound holds for all $U$-tree structures:
    \begin{equation}
        \lF{\star U} \le 2\sqrt{k}B.
    \end{equation}
\end{lemma}
\begin{proof}
    We have
    \begin{align}
        \lF{\star U}^2 
        &= \sum_{i=1}^k \ltwo{\star\uu_i}^2 \\
        &= \sum_{i=1}^k \ltwo{\star\w_i - \star\w_{\parent{i}}}^2 \\
        &\le \sum_{i=1}^k \left(\ltwo{\star\w_i} + \ltwo{\star\w_{\parent{i}}} \right)^2 \\
        &\le \sum_{i=1}^k (2B)^2 \\
        &\le k (2B)^2
        .
    \end{align}
\end{proof}
Now we consider the more important case of when we have a useful $U$-tree structure that meaningfully captures the similarity between classes.
This idea is captured in the following assumption.
\begin{assumption}
    \label{ass:metric}
    Let $\lambda >= 1$, and $d$ be a distance metric over the labels such that for all labels $i$ and $j$,
\begin{equation}
    \frac 1 \lambda d(i,j)
    \le \ltwo{\star \w_i - \star \w_j}
    \le \lambda d(i, j).
\end{equation}
We denote by $c$ the doubling dimension of the metric $d$,
and we assume that the $U$-tree structure is built using a cover tree \citep{beygelzimer2006cover}.
\end{assumption}

The Cover Tree was originally designed for speeding up nearest neighbor queries in arbitrary metric spaces.
The definition is rather technical, so we do not restate it here.
Instead, we mention only two important properties of the cover tree.
First, it can be constructed in time $O(k)$,
so building the cover tree is a cheap operation that has no meaningful impact on training performance.
It also only has to be done in the initialization step of the learning process, and not on every iteration.
Second, the cover tree construction has a hyperparameter which we call \texttt{base}.
This hyperparameter controls the fanout and depth of the tree because at any node $i$ at depth $\depth{i}$ in the tree,
the cover tree maintains the invariant that $d(i, \parent{i}) \le \texttt{base}^{-\depth{i}}$.
Increasing the $\texttt{base}$ hyperparameter results in shorter, fatter trees,
and decreasing it results in taller, narrower trees.
Our analysis follows the convention of setting $\texttt{base}=2$,
but we show in the experiments below that good performance is achieved for a wide range of base values.

The following Lemma uses Assumption \ref{ass:metric} and properties of the cover tree to bound the norm of $\lF{\star U}$.
It is our main technical result and the proof is deferred to the appendix.

\begin{lemma}
    \label{lemma:main}
    Under Assumptions \ref{ass:B} and \ref{ass:metric},
    when $c\le1$, we have that
    \begin{equation}
        \lF{\star U} \le \tfrac{1}{\sqrt2}\lambda B \sqrt{\log_2 k},
        \label{eq:c<=1}
    \end{equation}
    and when $c>1$, we have that
    \begin{equation}
        \lF{\star U} \le \sqrt{5}\lambda B \sqrt{k^{(1-1/c)}}.
        \label{eq:c>1}
    \end{equation}
\end{lemma}

We note that embedding techniques can be used to reduce the intrinsic dimension of the metric ($c$) at the expense of increasing the metric's distortion,
but we make no attempt to optimize this tradeoff in this paper.

We now state our main result.
It is an immediate consequence of the fact that $\lF{\star V} \le \lF{\star U}$ and Lemmas \ref{ref:cor:A} and \ref{lemma:main}.
\begin{corollary}
    \label{cor:main}
    Under assumptions \ref{ass:lip}, \ref{ass:B}, and \ref{ass:metric},
    when $c\le1$, the generalization error of the tree loss is bounded by
\begin{equation}
    \E L_D(\bar V) - L_D(V^*)
    \le \frac {\lambda B\rho \sqrt{\log_2 k}}{\sqrt 2n}
    .
\end{equation}
    When $c>1$, the generalization error of the tree loss is bounded by
\begin{equation}
    \E L_D(\bar V) - L_D(V^*)
    \le \frac {\lambda B\rho \sqrt{5 k^{(1-1/c)}}}{\sqrt n}
    .
\end{equation}
\end{corollary}
These convergence rates are asymptotically better than the convergence rates for the standard parameterization of the cross entropy loss.

@@

}

\clearpage
\appendix

It is well known that when using the standard cross entropy loss,
SGD converges at a rate of $O(\sqrt{k/n})$.
The tree loss's parameter sharing scheme asymptotically improves this convergence rate.
